Secure Composition of Robust and Optimising Compilers

### 2023-07-08 popl/main.tex
comments on 2.1 and 2.2

This paper employs the security model where programs are written in a language whose semantics emits events $\event$.
Events include security--relevant actions (e.g., reading from and writing to memory) and the unobservable event $\emptyevent$.
As programs execute, their emitted events are concatenated in traces $\trace$, which serve as the description of the behaviour of a program.%
\footnote{
Throughout the paper, we denote empty sequences with $\hole{\cdot}$ and concatenate sequences $\trace[_{1}],\trace[_{2}]$ as $\trace[_{1}]\cdot\trace[_{2}]$.
Prepending elements to sequences uses the same notation: $\event\cdot\trace$.
}

@@

The set of all properties can be partitioned into different {\em classes} ($\class$), i.e., safety, liveness, and neither safety or liveness.\MP{cit}
Given the central role for security, this paper focusses on safety properties.
Note that there are other classes $\class$ that contain more domain-specific properties, such as the class of memory safety properties.

@@

As an example, consider a trace that starts with an interaction with memory where a deallocation precedes a read: $\ev{Dealloc\ \loc}\cdot\ev{Read\ \loc\ 1729}\cdot\dots$
This program behaviour is insecure, because it reads from a memory location that was freed already.
The finite trace prefix $\ev{Dealloc\ \loc}\cdot\ev{Read\ \loc\ 1729}$ is enough to decide that the trace does not satisfy the property ,,no use after dealloc'' and there is no way to append events to this prefix which would result in the trace being admissible.

@@

%Given a whole program $\progvar$, i.e., a program without free variables, the {\em behaviour} ($\behav{\progvar}$) is the set containing all traces the program can emit during execution according to its language semantics.
%This paper only considers deterministic languages, where the behavior $\behav{\runtimetermvar}$ is a singleton set for any whole $\runtimetermvar$.
%This set is used to define property satisfaction (\Cref{def:propsat}):
\bul{Programs $\progvar$ satisfy a property $\pi$} iff \iul{$\progvar$ yields a trace $\trace$} such that \oul{$\trace$ is admissible with respect to $\pi$}.

\begin{definition}[Property Satisfaction]\label{def:propsat}
  $\progvar$ satisfies $\pi$, \bul{$\sat{\progvar}{\pi}$}, iff \iul{running $\progvar$ gives $\trace$} where \oul{$\trace\in\pi$}.
\end{definition}

Given a partial program $\progvar$, i.e., a program with free variables, the execution of $\progvar$ would get stuck due to the lack of definitions.
%As described earlier, partial programs cannot be executed \MP{not true anymore, need to define the notion of partial programs} , because they lack definitions and, thus, get stuck.
To remedy this, {\em linking} takes two partial programs $\progvar[_{1}],\progvar[_{2}]$ and produces a new whole program $\link{\progvar[_{1}]}{\progvar[_{2}]}$.
%Note that linking is not defined on programs, but merely on their syntactic descriptions, since programs contain dynamic state.\MP{ not terminology i really want to use, nor is defined}
As typically done in secure compilation~\cite{abate2019jour}, \MP{others do this too, parametricity results e.g.} this paper assumes that programs are the result of linking a {\em context} ($\ctx$) and a {\em component} ($\comp$).
% \MP{just introduce the notion of contexts}
%       MK: why only contexts?
%The component part of a whole program is the main object of interest and thus a known syntactic description.
The context is usually arbitrary\footnote{Up to, e.g., provoking link failure by giving multiple definitiions for the same name.} and thus has the role of an {\em attacker} that can interact with the component by means of whatever features the programming language has.
The semantics of the programming language is expected to differentiate between the component and the context.
%\MP{and say the semantics is expected (passive, hah) to differentiate between prg and ctx}
%%                                         -> why passive?
So, the context interacts in some way with the component and this interaction models an attacker.
With this, \Thmref{def:propsat} can be extended as follows.
For \bul{programs $\progvar$ to robustly satisfy a property $\pi$}, take an \iul{attacker $\contextvar$ and link it with $\progvar$}, \oul{the result of which must satisfy $\pi$}.
%The following definition extends on \Thmref{def:propsat} to cover property satisfaction under the presence of an (active, with respect to the semantics of the programming language) attacker: %, which is known as {\em robust satisfaction}:
%\MPin{missing: the context is an attacker! now, unlike before, we security property satisfaction with a clear active attacker}

\begin{definition}[Robust Satisfaction]\label{def:proprsat}
  $\progvar$ robustly satisfies $\pi$, \bul{$\rsat{\progvar}{\pi}$}, iff for \iul{any partial program $\contextvar$} it holds that \oul{$\sat{\link{\contextvar}{\runtimetermvar}}{\pi}$}.
\end{definition}

@@

A {\em compiler} ($\cc{\src{L}}{\trg{L}}$) translates syntactic descriptions of programs from a {\em source} ($\src{L}$) into a {\em target} ($\trg{L}$) programming language.
%For the syntactic description of programs written in $\src{L}$, this paper writes $\src{\progvar}$, while for ones written in $\trg{L}$, this paper writes $\trg{\progvar}$.%\MP{not a fan of this sentence}
%This paper writes $\src{\progvar}$ and $\trg{\progvar}$ for programs written in $\src{L}$ and $\trg{L}$, respectively.
%      MK: I don't think we need to mention this at all
This translation is considered {\em correct} if it is semantics--preserving.
%That is, the compiler encodes any possible $\src{L}$--level feature in $\trg{L}$, such that the semantics are the same.
Unfortunately, correct compilers may be insecure compilers\MK{citation needed}.
Programs translated by insecure compilers can violate security properties that the programmer assumes to hold. %, given the presence of a context--level program.
For a compiler to be secure, this paper makes use of the robust compilation framework~\cite{abate2019jour}, which we summarize in the following definition.
For \bul{compilers $\cc{\src{L}}{\trg{L}}$ to robustly preserve a class of properties $\class$}, consider \rul{properties $\pi$ of that class $\class$ and programs $\src{p}$ written in $\src{L}$} such that \iul{the program $\src{\progvar}$ robustly satisfies $\pi$}.
Then, \oul{$\cc{\src{L}}{\trg{L}}\left(\src{p}\right)$, the compiled version of $\src{\progvar}$, must robustly satisfy $\pi$}.

\begin{definition}[Robust Preservation]\label{def:rtp}
  Compiler $\cc{\src{L}}{\trg{L}}$ robustly preserves $\class$, \bul{$\rtp{\cc{\src{L}}{\trg{L}}}{\class}$}, iff \rul{for any $\pi\in\class$ and $\src{p}$} such that \iul{$\rsat{\src{\progvar}}{\pi}$}, it holds that \oul{$\rsat{\cc{\src{L}}{\trg{L}}\left(\src{p}\right)}{\pi}$}.
\end{definition}

@@

This section compares existing work with the one presented in this paper.
To this end, work on robust preservation (\Cref{subsec:relw:seccomprtp}) and on other secure compiler criteria (\Cref{subsec:relw:seccompcrit}) is looked at.
The case study of this paper (\Cref{sec:casestud:defs,sec:casestud:rtp}) implements measures for ensuring \gls{ms} as well as \gls{cct} and these are compared with previous work as well (\Cref{subsec:relw:msmechs,subsec:relw:cctmechs}).

@@

The robust preservation of properties as a compiler--level criterion has been analyzed extensively~\cite{abate2019jour,patrignani2021rsc,abate2021extacc,patrignani2022universal,patrignani2019survey,kruse2022csc}.
This work has made use of that previous work.
While there exists a composition theorem~\cite{abate2019jour} already, that theorem is not concerned with composing robustly safe compilers, but rather program components and contexts.
The work relating robust preservation with universal composability~\cite{patrignani2022universal} is closest to what this paper presents.
The authors demonstrate a similar compositionality theorem to what is presented here (\Cref{sec:compcomp}) as well as in an earlier version of this work~\cite{kruse2022csc}.
However, they do not demonstrate the scalability of the approach by means of a case study.

@@

Different mechanisms for memory safety exist that also consider the secure compilation domain, i.e., have an active attacker model.
For example, the ,,pointers as capabilities'' principle represents pointers as machine--level capabilities~\cite{korashy2021capableptrs}, which behave in a similar fashion to capabilities by means of linear typing~\cite{morrisett2005L3}.
The approach of this paper also uses linear typing, but differs from $L^{3}$~\cite{morrisett2005L3} in the way that functions are not first--class.
Moreover, this paper considers an active attacker, while the work on $L^{3}$ only discusses whole programs and, thus, has no active attacker model.
The instrumentation to ensure memory safety that this paper presents is inspired by seminal work in the literature~\cite{nagarakatte2009soft}.
That work inserts bounds--checks in front of pointer--dereferences and, for this to work, inserts meta--data information on pointer creation.
It also works in a more advanced setting with structured fields accesses and also introduces a table--lookup for pointers that are stored in memory.
This paper only considers pointers to arrays of simple objects, i.e., there are no pointers to pointers or structures.
Of course, there are a lot of other approaches to memory--safety in literature, specifically as compiler instrumentations~\cite{akritidis2009baggy,younan2010paricheck,jung2021pico,shankaranarayana2023tailcheck,dhumbumroong2020boundwarden,nam2019framer,zhou2023fatptrs}, hardware--extensions~\cite{kwon2013lowfat,saileshwar2022heapcheck,chen2023flexpointer,kim2023whistle}, or programming language extensions~\cite{elliott2018checkedc,li2022formalcheckedc,jim2002cyclone,elliott2015guilt,west2005cuckoo,weis2019fyr,benoit2019uniqueness}, to name a few.
What differentiates this work from them is that this work uses known, compiler--based approaches to ensure memory--safety as a means to investigate secure compiler compositions.
This paper does not provide efficient memory--safety, but serves as a theoretical foundation for the secure compilation domain.

To extend the languages in this paper with a less restricted form of pointer arithmetic, the region coloring memory safety monitor presented in earlier work~\cite{michael2023mswasm} can be used.
The work presenting this monitor provides an approach for the robust preservation of memory safety compiling from C to WASM.
However, they do not discuss composition of secure compilers but rather investigate an instance of a secure compiler.

@@

The approach to cryptographic constant time in this paper is high--level, where a programming language exposes a way to switch the semantics to a data (operand) independent timing mode.
Since identifiers in $\Lscct$ are annotated with a secrecy tag, this approach is similar to others with information flow control.
For example, Vale~\cite{bond2017vale} uses Dafny to ensure constant--time assembly code, while Jasmin~\cite{almeida2017jasmin} makes use of the Coq proof assistant to reject non--constant--time programs.
CT-Wasm~\cite{watt2019ctwasm} enforces constant--timeness by means of a type system.
Different to the approach of this paper, these approaches necessitate that the programmer writes \gls{cct} code.
An approach to allow programmers to write more high--level code is CryptOpt~\cite{kuepper2023cryptopt}, which generates efficient target-code by means of a randomized search.
This paper abstracts over concrete mitigation strategies and simply assumes that there is a flag to switch to a cryptographic--constant time execution mode.
This can be realized by employing the FaCT~\cite{cauligi2019fact} compiler, which translates common non-constant time code patterns to be constant--time, and the data (object) independent timing execution mode of modern processors.

>>>>

This paper employs the security model where programs are written in a language whose semantics emits events $\event$.
Events include security--relevant actions (e.g., reading from and writing to memory) and the unobservable event $\emptyevent$.
As programs execute, their emitted events are concatenated in traces $\trace$, which serve as the description of the behaviour of a program.%
\footnote{
Throughout the paper, we indicate sequences with an overbar (i.e., $\trace$), we denote empty sequences with $\hole{\cdot}$ and concatenate sequences $\trace[_{1}],\trace[_{2}]$ as $\trace[_{1}]\cdot\trace[_{2}]$.
Prepending elements to sequences uses the same notation: $\event\cdot\trace$.
}

@@

The set of all properties can be partitioned into different {\em classes} ($\class$), i.e., safety, liveness, and neither safety nor liveness.\MP{cit}
Given their central role for security, this paper focusses on safety properties.
% Note that there are other classes $\class$ that contain more domain-specific properties, such as the class of memory safety properties.
\MPin{
	remove.
	what does this sentence add? removing it does not remove any relevant information

@@

\MPin{
	passive. also, ``a trace admits a property'' is wrong, it's the other way around
As an example, consider a trace describing an interaction with a memory where the deallocation of a pointer $\loc$ precedes its read: $\ev{Dealloc\ \loc}\cdot\ev{Read\ \loc\ 1729}\cdot\dots$
This program behaviour is insecure wrt a canonical notion of (temporal) memory safety dictating no use-after-frees of pointers\MP{cit}, because it reads from a memory location that was freed already.
\MPin{
	rewordeda a little for precision (e.g., what is $\loc?$), check the diff
The previous finite trace prefix is enough to decide that the trace does not satisfy temporal memory safety and there is no way to append events to this prefix which would result in the trace being admissible.

@@

%Given a whole program $\progvar$, i.e., a program without free variables, the {\em behaviour} ($\behav{\progvar}$) is the set containing all traces the program can emit during execution according to its language semantics.
%This paper only considers deterministic languages, where the behavior $\behav{\runtimetermvar}$ is a singleton set for any whole $\runtimetermvar$.
%This set is used to define property satisfaction (\Cref{def:propsat}):
\MPin{
	fill the blank of the next sentence.
}
In the following we write that a whole program \MP{} runs according to the language semantics and produces trace \MP{} as \MP{}. 
With this, we define that
\bul{programs $\progvar$ satisfy a property $\pi$} iff \iul{$\progvar$ yields a trace $\trace$} such that \oul{$\trace$ is admissible with respect to $\pi$} (\Cref{def:propsat}).

\begin{definition}[Property Satisfaction]\label{def:propsat}
  % $\progvar$ satisfies $\pi$, 
  \bul{$\sat{\progvar}{\pi}$} 
  % iff 
  $\isdef$
  \iul{running $\progvar$ gives $\trace$}
  \MP{this does not formalise things, see sentence added above and add formalism}
   where \oul{$\trace\in\pi$}.
\end{definition}

\MPin{
	why are we suddenly going from talking about prop sat to talknig about progs and ctxs?
	there's a gap here that needs to be bridged.
	A plausible solution is something along these lines.
	"property satisfaction only considers passive attackers, to consider a stronger model with active attackers existing work (as well as this paper) uses robust property satisfaction, which we define after introducing the required concept of program context and partial programs"
}
Given a partial program $\progvar$, i.e., a program with free variables, the execution of $\progvar$ would get stuck due to the lack of program parts (e.g., function definitions).
%As described earlier, partial programs cannot be executed \MP{not true anymore, need to define the notion of partial programs} , because they lack definitions and, thus, get stuck.
To remedy this, {\em linking} takes two partial programs $\progvar[_{1}],\progvar[_{2}]$ and produces a new whole program $\link{\progvar[_{1}]}{\progvar[_{2}]}$.
\MPin{
	cumbersome notation.
	`link ... ... ' should be used to identify the notation of the operation, but we should have a metavariable for the result, like `link(p1,p2) = w'
	W is a known candidate
}
%Note that linking is not defined on programs, but merely on their syntactic descriptions, since programs contain dynamic state.\MP{ not terminology i really want to use, nor is defined}
As typically done in secure compilation~\cite{abate2019jour}, \MP{others do this too, parametricity results e.g.} this paper assumes that \MP{whole?}programs are the result of linking a {\em context} ($\ctx$) and a {\em component} ($\comp$).
\MPin{
	these sentences identify the need to have 2 different terms: 1 for partial programs (components?) one for whole programs (just programs) and to use them consistently.
}
% \MP{just introduce the notion of contexts}
%       MK: why only contexts?
%The component part of a whole program is the main object of interest and thus a known syntactic description.
The context is an arbitrary program and thus has the role of an {\em attacker} that can interact with the component by means of whatever features the programming language has.
In this work, the semantics of the programming language is expected to differentiate between the component and the context.
\MPin{
	so what?
}
%\MP{and say the semantics is expected (passive, hah) to differentiate between prg and ctx}
%%                                         -> why passive?
So, the context interacts in some way with the component and this interaction models an attacker.
\MP{
	wut?
	needless sentence, erase.
}
With this, \Thmref{def:propsat} can be extended as follows: for \bul{ partialprograms $\progvar$ to robustly satisfy a property $\pi$}, take an \iul{attacker $\contextvar$ and link it with $\progvar$}, \oul{the result of which must satisfy $\pi$}.
\MPin{
	the behaviour of the result?
}
%The following definition extends on \Thmref{def:propsat} to cover property satisfaction under the presence of an (active, with respect to the semantics of the programming language) attacker: %, which is known as {\em robust satisfaction}:
%\MPin{missing: the context is an attacker! now, unlike before, we security property satisfaction with a clear active attacker}

\begin{definition}[Robust Satisfaction]\label{def:proprsat}
  % $\progvar$ robustly satisfies $\pi$, 
  \bul{$\rsat{\progvar}{\pi}$}
  % iff 
  $\isdef$ for \iul{any partial program $\contextvar$}\MP{use $\forall$} it holds that \oul{$\sat{\ {\contextvar}{\runtimetermvar}}{\pi}$}. \MP{why r?}
\end{definition}

@@

A {\em compiler} ($\cc{\src{L}}{\trg{L}}$) translates syntactic descriptions of programs from a {\em source} ($\src{L}$) into a {\em target} ($\trg{L}$) programming language.
%For the syntactic description of programs written in $\src{L}$, this paper writes $\src{\progvar}$, while for ones written in $\trg{L}$, this paper writes $\trg{\progvar}$.%\MP{not a fan of this sentence}
%This paper writes $\src{\progvar}$ and $\trg{\progvar}$ for programs written in $\src{L}$ and $\trg{L}$, respectively.
%      MK: I don't think we need to mention this at all
This translation is considered {\em correct} if it is semantics--preserving.
\MPin{
	i.e., ... brief desc?
}
%That is, the compiler encodes any possible $\src{L}$--level feature in $\trg{L}$, such that the semantics are the same.
Unfortunately, correct compilers may be insecure compilers\MK{citation needed}, and programs translated by insecure compilers can violate security properties that the programmer assumes to hold. %, given the presence of a context--level program.
To define when a compiler is secure, this paper uses the robust compilation framework~\cite{abate2019jour}, which we summarize in the following definition.

To define robust compilation we only need the notion of a class of properties.
\MPin{define.}

For \bul{compilers $\cc{\src{L}}{\trg{L}}$ to robustly preserve a class of properties $\class$}, if for any \rul{property $\pi$ of that class $\class$ and programs $\src{p}$ written in $\src{L}$} we have that \iul{the program $\src{\progvar}$ robustly satisfies $\pi$}, then \oul{the compilation of $\src{\progvar}$, $\cc{\src{L}}{\trg{L}}\left(\src{p}\right)$, must robustly satisfy $\pi$}.

\begin{definition}[Robust Preservation]\label{def:rtp}
  % Compiler $\cc{\src{L}}{\trg{L}}$ robustly preserves $\class$, 
  \bul{$\rtp{\cc{\src{L}}{\trg{L}}}{\class}$}
  %, iff 
  $\isdef$
  \rul{for any $\pi\in\class$ and $\src{p}$} such that \iul{$\rsat{\src{\progvar}}{\pi}$}\MP{ use $\forall$}, it holds that \oul{$\rsat{\cc{\src{L}}{\trg{L}}\left(\src{p}\right)}{\pi}$}.
\end{definition}

@@

This section compares existing work with the one presented in this paper.
To this end, work on robust preservation criteria (\Cref{subsec:relw:seccomprtp}) and on other secure compiler criteria (\Cref{subsec:relw:seccompcrit}) is being looked at.
The case study of this paper (\Cref{sec:casestud:defs,sec:casestud:rtp}) implements measures for ensuring \gls{ms} as well as \gls{cct} and these are compared with previous work as well (\Cref{subsec:relw:msmechs,subsec:relw:cctmechs}).

@@

The robust preservation of properties as a compiler--level criterion has been analyzed extensively~\cite{abate2019jour,patrignani2021rsc,abate2021extacc,patrignani2022universal,patrignani2019survey,kruse2022csc}.
This work has made use of that previous work.
While there exists a composition theorem~\cite{abate2019jour} already, that theorem is not concerned with composing robustly safe compilers, but rather program components and contexts.
The work relating robust preservation with universal composability~\cite{patrignani2022universal} is closest to what this paper presents.
The authors demonstrate a similar compositionality theorem to what is presented here (\Cref{sec:compcomp}) as well as in an earlier version of this work~\cite{kruse2022csc}.
However, they do not demonstrate the scalability of the approach in the context of a compilation chain.

@@

There is still no consensus what the formal definition of memory safety as a program property should be. %cite
Yet, different mechanisms for memory safety exist that also consider the secure compilation domain, i.e., have an attacker model that is not just the observing attacker.
For example, the ,,pointers as capabilities'' principle represents pointers as machine--level capabilities~\cite{korashy2021capableptrs}, which behave in a similar fashion to capabilities by means of  linear typing~\cite{morrisett2005L3}.
The approach of this paper to memory safety and the countermeasures against memory safety insecurities are inspired by seminal work in the literature~\cite{nagarakatte2009soft,nagarakatte2010cets}.
The view of memory safety in this paper exhibits the classic split between temporal and spatial memory safety.
The former making sure that there are no double deallocations or use after frees.
The latter making sure that there are no out of bounds accesses.
When extending the languages in this paper with a less restricted form of pointer arithmetic, the region coloring memory safety monitor presented in earlier work~\cite{michael2023mswasm} can be used.
The work presenting this monitor provides an approach for the robust {\em enforcement}, which implies robust preservation, of memory safety.
But, they do not compose the enforcement mechanism with other secure compilation passes.

@@
